Sampling Bayesian probabilities given only sampled priors

### 2025-05-23 main.tex
post-CWR updates including author list, aastex7

% section/paper on sampling with sampled priors.
% \documentclass[preprint,linenumbers]{aastex631} # 6.31 fails with \align!!
\documentclass[linenumbers, onecolumn]{aastex63}
\usepackage{xcolor}

@@

\author[0000-0002-8613-8259]{Gary M. Bernstein}
\affiliation{Department of Physics and Astronomy, University of Pennsylvania, Philadelphia, PA 19104, USA}
\email{garyb@upenn.edu}

\author{William d'Assignies}
\author{Michael Troxel}
\author{Alex Alarcon}
\author{Alex Amon}
\author{Giulia Giannini}
\author{Boyan Yin}
\author{Builders\ldots} %%%%

\begin{abstract}
	\vspace{0.2in}
A typical Bayesian inference on the values of some parameters of
interest $\vecq$ from some data $D$ would run a Markov Chain (MC) to sample
from the posterior $p(\vecq,\vecn | D) \propto \likeli(D | \vecq,\vecn)
p(\vecq) p(\vecn),$ where $\vecn$ are some nuisance parameters.  In
many cases, the nuisance parameters are high-dimensional, and their
prior $p(\vecn)$ is itself defined only by a set of samples that have
been drawn from some other MC.  Two problems arise: first,
the MC for the posterior will typically require evaluation of
$p(\vecn)$ at arbitrary values of $\vecn,$ \ie\ one needs to provide a
density estimator over the full $\vecn$ space from the provided
samples.  Second, the high dimensionality of $\vecn$ hinders both the
density estimation and the efficiency of the MC for the posterior.  We
describe a solution to this problem: a linear compression of the
$\vecn$ space into a much lower-dimensional space $\vecu$ which
projects away directions in $\vecn$ space that cannot appreciably
alter $\likeli.$ The algorithm for doing so is a slight modification
to principal components analysis, and is less restrictive on
$p(\vecn)$ than other proposed solutions to this issue.
We demonstrate this ``mode projection'' technique using the analysis
of 2-point correlation functions of weak lensing fields and galaxy
density in the \textit{Dark Energy Survey}, where $\vecn$ is a binned representation of the redshift distrbution
$n(z)$ of the galaxies.
\end{abstract}
\reportnum{}

@@

The scenario we address here is when \emph{the prior $p(\vecn)$ is
  itself known only from a set of samples of $\vecn$ from this
  distribution.} Most MC samplers require that the posterior be an
evaluable function of any value of the parameters. It is the general task of density estimators to convert the samples of $\vecn$ into an evaluable $p(\vecn).$  But when \vecn\ is of high dimension, two problems arise: first, there may be insufficient available samples to create a viable density estimator; second, sampling of the posterior in (\ref{eq:posterior}) becomes infeasible if the MC must traverse a high-dimensional space.

@@

\section{Application}\label{sec:app}
As an example of the application of this straightforward dimensional
reduction to a high-dimensional nuisance parameter, we examine the
redshift distribution of one of the bins of ``Maglim''  galaxies used
as a lens population and clustering tracer in the Year 6 (Y6) analysis of
the DES galaxy catalogs.  Each of these Maglim ``lens bins''
is selected with cuts on galaxy fluxes and
colors in an attempt to generate a sample that is confined to a
particular redshift range, as described in \citet{y6gg}.  Once each bin's
sample is chosen, a combination of 
photometric techniques \citep{y6pz} and clustering information
\citep{y6wz} is used to generate samples from the posterior
distribtuion of  the $n(z)$ distribution of the bin members, conditioned
on the photometric and clustering data.

@@

@@

\begin{thebibliography}{dummy}
%\bibliography
%\bibliographystyle{aasjournal}
\bibitem[d'Assignies et al.(2025)]{y6wz} d'Assignies, W. et al.\ 2025
  (WZ paper) ???
\bibitem[Bridle et al.(2002)]{bridle02}
Bridle, S. L., Crittenden, R., Melchiorri, A., Hobson, M. P., Kneissl, R., and Lasenby, A. N. 2002, \mnras, 335, 1193, 
doi:10.1046/j.1365-8711.2002.05709.x

@@

\bibitem[Giannini et al.(2025)]{y6gg} Giannini, G. et al.\ 2025
  (lenses paper) ????

@@

\bibitem[Yin et al.(2025)]{y6pz} Yin, B. et al.\ 2025 (PZ paper) ???

@@

\bibitem[??? et al.(2025)]{y6model}
  Y6 modeling paper ???

@@

\end{document}

>>>>

% section/paper on sampling with sampled priors.
% \documentclass[preprint,linenumbers]{aastex631} # 6.31 fails with \align!!
\documentclass[linenumbers, onecolumn]{aastex7}
\usepackage{xcolor}

@@

\include{authors.tex}
% \author[0000-0002-8613-8259]{Gary M. Bernstein}
% \affiliation{Department of Physics and Astronomy, University of Pennsylvania, Philadelphia, PA 19104, USA}
% \email{garyb@upenn.edu}

% \author{William d'Assignies}
% \author{Michael Troxel}
% \author{Alex Alarcon}
% \author{Alex Amon}
% \author{Giulia Giannini}
% \author{Boyan Yin}
% \author{Builders\ldots} %%%%

\begin{abstract}
	\vspace{0.2in}
A typical Bayesian inference on the values of some parameters of
interest $\vecq$ from some data $D$ involves running a Markov Chain (MC) to sample
from the posterior $p(\vecq,\vecn | D) \propto \likeli(D | \vecq,\vecn)
p(\vecq) p(\vecn),$ where $\vecn$ are some nuisance parameters.  In
many cases, the nuisance parameters are high-dimensional, and their
prior $p(\vecn)$ is itself defined only by a set of samples that have
been drawn from some other MC.  Two problems arise: first,
the MC for the posterior will typically require evaluation of
$p(\vecn)$ at arbitrary values of $\vecn,$ \ie\ one needs to provide a
density estimator over the full $\vecn$ space from the provided
samples.  Second, the high dimensionality of $\vecn$ hinders both the
density estimation and the efficiency of the MC for the posterior.  We
describe a solution to this problem: a linear compression of the
$\vecn$ space into a much lower-dimensional space $\vecu$ which
projects away directions in $\vecn$ space that cannot appreciably
alter $\likeli.$ The algorithm for doing so is a slight modification
to principal components analysis, and is less restrictive on
$p(\vecn)$ than other proposed solutions to this issue.
We demonstrate this ``mode projection'' technique using the analysis
of 2-point correlation functions of weak lensing fields and galaxy
density in the \textit{Dark Energy Survey}, where $\vecn$ is a binned representation of the redshift distrbution
$n(z)$ of the galaxies.
\end{abstract}
\reportnum{}

@@

The scenario we address here is when \emph{the prior $p(\vecn)$ is
  itself known only from a set of samples of $\vecn$ from this
  distribution.} Most MC samplers require that the posterior (and
hence the prior and likelihood) be an
evaluable function of any value of the parameters. It is the general task of density estimators to convert the samples of $\vecn$ into an evaluable $p(\vecn).$  But when \vecn\ is of high dimension, two problems arise: first, there may be insufficient available samples to create a viable density estimator; second, sampling of the posterior in (\ref{eq:posterior}) becomes infeasible if the MC must traverse a high-dimensional space.

@@

\section{Application}\label{sec:app}
As an example of the application of this straightforward dimensional
reduction to a high-dimensional nuisance parameter, we examine the
redshift distribution of one of the bins of ``Maglim''  galaxies used
as a lens population and clustering tracer in the Year 6 (Y6) analysis of
the DES galaxy catalogs.  Each of these Maglim ``lens bins''
is selected with cuts on galaxy fluxes and
colors in an attempt to generate a sample that is confined to a
particular redshift range, as described in \citet{y6maglim}.  Once each bin's
sample is chosen, a combination of 
photometric techniques \citep{y6lenspz,y6pz} and clustering information
\citep{y6wz} is used to generate samples from the posterior
distribtuion of  the $n(z)$ distribution of the bin members, conditioned
on the photometric and clustering data.

@@

For simplicity, we have demonstrated this method for a case in which
$\vecn$ specifies the $n(z)$ function for a single sample of DES
galaxies.  The method is fully applicable to any set of nuisance
parameters for which we are given a set of samples from $p(\vecn).$
For instance, in the DES Y6 analysis, we use a $\vecn$ that is a
concatenation of the parameters of $n(z)$ for 4 bins of lens source
galaxies.  Since each output sample from the  redshift-estimate
process describes all 4 bins, we know the cross-correlations between
the bins' $n(z)$ behavior, and the compression process described
herein will properly preserve these correlations in the subsequent
cosmological analysis.

@@

\begin{thebibliography}{dummy}
%\bibliography
%\bibliographystyle{aasjournal}
% Y6 WZ paper
\bibitem[d'Assignies et al.(2025)]{y6wz} d'Assignies, W. et al.\ 2025
  (in preparation)

\bibitem[Bridle et al.(2002)]{bridle02}
Bridle, S. L., Crittenden, R., Melchiorri, A., Hobson, M. P., Kneissl, R., and Lasenby, A. N. 2002, \mnras, 335, 1193, 
doi:10.1046/j.1365-8711.2002.05709.x

@@

% lenses paper
\bibitem[Giannini et al.(2025)]{y6lenspz} Giannini, G. et al.\ 2025 (in preparation)

@@

\bibitem[Giannini et al.(2025)]{y6maglim} Weaverdyck, N. et al.\ 2025 (in preparation)

\bibitem[Yin et al.(2025)]{y6pz} Yin, B. et al.\ 2025 (in preparation)

@@

% Y6 modelling paper
\bibitem[Sanchez-Cid et al.(2025)]{y6model} Sanchez-Cid et al.\ 2025
  (in preparation)

@@

\allauthors
\end{document}
